---
title: "From Cape to Cairo: Following African Economic Crises from the Colonial Era to through the Modern Age"
author: "Six-Thirty-Eight"
date: "10/22/19"
output: github_document
---

```{r load-packages, warning = F}
library(tidyverse)
```

### Section 1. Introduction

  For our STA 199 project, we chose to examine the stability and development of the economies of African nations, particularly in the context of colonialism and decolonialism. After we had read an article about Zimbabwe’s crippling inflation crisis (in which the nation completely abandoned paper currencies altogether), we wondered: how has Africa’s history of exploitation and unrest contributed to the continent’s overall economic instability in the modern day, and what factors be correlated with an economic crisis in the continent? Thus, we collected three datasets to answer this question and the questions that precipitated from it. The first of these datasets was an excellently curated “African Economic Banking and Systemic Crises Data” from a Kaggle (https://www.kaggle.com/chirin/africa-economic-banking-and-systemic-crisis-data/version/1#) and the global economic crisis dataset from which it was derived, created by the Harvard Business School (https://www.hbs.edu/behavioral-finance-and-financial-stability/data/Pages/global.aspx). While these datasets gave insights into instances of “systemic crisis” and exchange rate and CPI information (which indicates relative inflation and debt), they critically omitted the GDP of these nations over time – a fairly modern economic growth metric that we were only able to find being consistently used from 1960 onward. We sourced and cleaned a record of this global GDP data from the World Bank (https://data.worldbank.org/indicator/ny.gdp.mktp.cd), detailing world nations’ GDPs in USD from 1960 to the modern day. We believe that this dataset will be sufficient to detail the impacts of decolonization on African nations’ GDPs (as most African decolonization efforts took place in the mid-to-late 1950s and early 1960s) and provide a working conclusion to our research questions.

### Section 2. Exploratory data analysis

```{r load-data}
africa <- read_csv("../data/african_crises.csv")
global <- read_csv("../data/global_crisis_data.csv")
gdps <- read_csv("../data/world_gdp_data.csv")

gdps <- gdps %>%
  pivot_longer(
    cols = c(
      -`Country Name`,
      -`Country Code`,
      -`Indicator Name`,
      -`Indicator Code`),
    names_to = "year"
  ) %>%
  rename(cc3 = `Country Code`, country = `Country Name`, gdp = value)

africangdps <- gdps %>%
  merge(., africa, by = c("country","cc3","year"))

global <- global %>%
  select(-`<`)
```

To start, we can visualize the number of years with a banking crisis in African countries as a bar graph. We can also look at how African GDPs compare to nations in other continents to emphasize why our question of instability is important.

```{r banking-crisis}
africa %>%
ggplot(aes(x = banking_crisis)) +
  geom_bar(fill = "pink") +
  labs(
    x = "Banking Crisis", 
    title = "Number of Years with Banking Crises in African Countries"
    )
```

We can create a plot of all the countries included in the African economic dataset and what years data is given for them.

```{r countries-included, warning = F}
ggplot(data = africa, mapping = aes(x = year)) +
  geom_histogram(binwidth = 10, fill = "blue", col = "black") +
  labs(
    x = "Year",
    y = "Number of Countries", 
    title = "Increase in Total Number of African
    Country Economic Observations Over Time"
    )
ggplot(data = africa, mapping = aes(x = country, y = year)) +
  geom_point(size = .75, col = "orange") +
  coord_flip() +
  labs(
  x = "Country",
  y = "Year",
  title = "African Economic Data Observed Each Year by Country"
  )
```

Overall, observations of African economies greatly increased after 1950. However, there are some African countries, such as Egypt and South Africa, whose economies have been monitored for over a century.

Next, we can determine the percentage of systemic crisis years for each country in the Africa dataset.

```{r systemic-crises}
africa %>%
  count(country, systemic_crisis) %>%
  group_by(country) %>%
  mutate(crisis_proportion = 1- (n/sum(n))) %>%
  filter(systemic_crisis == 0) %>%
  select(country, crisis_proportion) %>%
  arrange(desc(crisis_proportion))
```

The Central African Republic had the highest proportion of recorded years with systemic crises, with 32.8% of the years recorded having a systemic crisis. Out of the 13 African countries in the dataset, only South Africa, Angola, and Mauritius had no systemic crises.

Now, we'll explore an interesting question. Independence movements are tricky - while they might achieve democracy and self-governance for a country, they often contribute to political and economic instability. For each country, we'll determine how many years after independence a country will typically experience it's next crisis.

```{r independence, warning = F}
output <- tibble(country = distinct(africa, country)$country)
output$independence_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(row_number()==1) %>%
  ungroup() %>%
  select(year)
  
output$crisis_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(banking_crisis == "crisis") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(year)
output <- output %>%
  mutate(difference = (crisis_year$year - independence_year$year))
ggplot(data = output, mapping = aes(y = difference)) +
  geom_boxplot() + 
  labs(
    title = "What's the typical amount of years between a country 
    achieving independence and its next financial crisis?",
    y = "Number of Years")
output %>%
  summarise(IQR = IQR(difference), median = median(difference))
```

We see that the median amount of years a country will first encounter a banking crisis after they achieve independence is about 30 years, with an interquartile range of 11 years. This brings us to an interesting question - what values characterize countries that had a long period of economic prosperity after independence, and what characterizes countries that saw immediate financial crises? 

Next, we'll do a univariate visualization of what the exchange rate variable looks like:
```{r exchange-rate-visualize}
ggplot(data = africa, mapping = aes(x = exch_usd)) +
  geom_histogram(bins = 10)

summary_stats <- africa %>%
  summarise(median = median(exch_usd), std = sd(exch_usd))
```
We see that the median exchange rate against the US dollar is `r summary_stats$median` and the standard deviation is `r summary_stats$std`. The distribution is right-skewed. 

Next, let's analyze if exchange rate against the dollar and inflation rate correlate with economic crises.

```{r inflation-crisis-relationship, warning = F}
africa %>%
ggplot(
  aes(
    x = inflation_annual_cpi,
    y = exch_usd,
    color = banking_crisis)) +
  geom_point() + 
  xlim(-10, 100)
```

We see that high levels of inflation often correspond to banking crises, but this is not always true. There were many cases of banking crises that existed even when inflation was lower than typical. Moreover, banking crises also tended to exist when exchange rates against the dollar were high. This could point to a pretty interesting research question: what levels of inflation and exchange rate against the dollar will signal a banking crisis (or any crisis in general?)

### Section 3. Research questions

Questions:

1. How did major European economies change with colonialism? How did African economies change when colonized? Conversely, how did European/African economies change with decolonization of the African continent?

To answer this question, we will look at changes in GDP immediately following colonization and decolonization for both the colonizing country and the colonized country. Naturally, we will be using `gdp` from our gdps dataframe (see environment) as a response variable for the predictor variables `independence` and `year`. We may also incorporate `Systemic Crisis` from the `global` dataframe to see whether a nation immediately colonized or immediately acquiring a new colony suffered economic instability, perhaps as a result of the event.

2. Was an economic crisis more likely following n years after decolonization?

To answer this question, we will examine the number of years it takes for a country to enter a systemic, economic, banking, or inflation crisis following independence, and which of these crises is more common for fledgling independent nations. For countries who have not faced crisis following decolonization, we will observe what factors of their economy contribute to their relative stability. As a preliminary data manipulation in Section 2 has shown, we will be using the `crisis` variables from the `africa` dataset as response variables for a new variable `difference`, which takes the difference between the year of the first crisis and the year of independence. To understand other economic variables that contribute to stability, see below.

3. What factors are most associated with a systemic crisis in Africa? How can identify a systemic crisis in the dataset? How do African systemic crises compare to other nations' crises? 

To answer this question, we will make a regression model or simulation relating systemic crises to the various variables in our dataset, and optimize for the highest adjusted r-squared value. Thus, `Systemic Crises` from the `global` dataset filtered for African nations will be a response variable to changes in as-of-yet unknown predictor variables (our goal is to find these variables). After discovering these variables for African nations, we will filter out African nations and reoptimize the model. Our goal is to see what variables predict systemic crises in developed nations and how these variables differ (either whether they are indeed the same variables or the r-squared impact they have on the model) from the variables in the African-only model.

### Section 4. Data

```{r glimpse-africa-data, warning = F, message = F}

glimpse(africa)
glimpse(global)
glimpse(gdps)

```