---
title: "From Cape to Cairo: Following African Economic Crises from the Colonial Era to through the Modern Age"
author: "Six-Thirty-Eight"
date: "10/22/19"
output: github_document
---

```{r load-packages, warning = F}
library(tidyverse)
```

### Section 1. Introduction

Beck

### Section 2. Exploratory data analysis

Isabelle and Jerry

### Section 3. Research questions

Questions:

1. How did major European economies change with colonialism? How did African economies change when colonized? Conversely, how did European/African economies change with decolonization of the African continent?

 - Danger: there are a lot of factors that affect a European economies, and losing a colony might not be that impactful. Concentrate on African economies in this analysis.
 
2. Was an economic crisis more likely following n years after decolonization?

3. What factors are most associated with a systemic crisis in Africa? How can identify a systemic crisis in the dataset? What defines a systemic crisis, and how do African systemic crises compare to other nations' crises? 

### Section 4. Data

```{r glimpse-africa-data, warning = F, message = F}

africa <- read_csv("/cloud/project/data/african_crises.csv")
global <- read_csv("/cloud/project/data/global_crisis_data.csv")

glimpse(africa)

glimpse(global)
```

