---
title: "From Cape to Cairo: Following African Economic Crises from the Colonial Era to through the Modern Age"
author: "Six-Thirty-Eight"
date: "10/22/19"
output: github_document
---

```{r load-packages, warning = F}
library(tidyverse)
```

### Section 1. Introduction

Beck

### Section 2. Exploratory data analysis
Isabella and Jerry

```{r load-data}
africa <- read_csv("../data/african_crises.csv")
```

```{r independence}

output <- tibble(country = distinct(africa, country)$country)

output$independence_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(row_number()==1) %>%
  ungroup() %>%
  select(year)
  

output$crisis_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(banking_crisis == "crisis") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(year)

output <- output %>%
  mutate(difference = (crisis_year$year - independence_year$year))

ggplot(data = output, mapping = aes(y = difference)) +
  geom_boxplot()

output %>%
  summarise(IQR = IQR(difference), median = median(difference))
```
We see that the median amount of years a country will first encounter a banking crisis after they achieve independence is about 30 years, with an interquartile range of 11 years. This brings us to an interesting question - what values characterize countries that had a long period of economic prosperity after independence, and what characterizes countries that saw immediate financial crises? 

```{r inflation-crisis-relationship}
ggplot(data = africa, mapping = aes(x = inflation_annual_cpi, y = exch_usd, color = banking_crisis, alpha = 0.5)) +
  geom_point() + 
  xlim(-50, 100)

ggplot(data = africa, mapping = aes(x = inflation_annual_cpi, y = exch_usd, color = banking_crisis)) +
  geom_point() + 
  xlim(-10, 100)

```
We see that high levels of inflation often correspond to banking crises - but this is not always true. There were many cases of banking crises that existed even when inflation was lower than typical. Moreover, banking crises also tended to exist when exchange rates against the dollar were high. This could point to a pretty interesting research question - what levels of inflation and exchange rate against the dollar will signal a banking crisis (or any crisis in general?)

### Section 3. Research questions

Questions:

1. How did major European economies change with colonialism? How did African economies change when colonized? Conversely, how did European/African economies change with decolonization of the African continent?

 - Danger: there are a lot of factors that affect a European economies, and losing a colony might not be that impactful. Concentrate on African economies in this analysis.
 
2. Was an economic crisis more likely following n years after decolonization?

3. What factors are most associated with a systemic crisis in Africa? How can identify a systemic crisis in the dataset? What defines a systemic crisis, and how do African systemic crises compare to other nations' crises? 

### Section 4. Data

```{r glimpse-africa-data, warning = F, message = F}

africa <- read_csv("/cloud/project/data/african_crises.csv")
global <- read_csv("/cloud/project/data/global_crisis_data.csv")

glimpse(africa)

glimpse(global)
```

