---
title: "An Analysis of African Systemic Crises"
author: "Beck Addison, Jerry Lin, Isabella Swigart, Emma Hirschkop"
date: "12/13/2019"
output: 
  github_document:
      pandoc_args: --webtex
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

## Introduction
```{r load-packages, message = F}

#install.packages(c("infer","countrycode","gridExtra"))
library(infer)
library(tidyverse)
library(countrycode)
library(broom)
library(knitr)
library(gridExtra)
```

```{r load-data-and-initialization, message = F, warning = F}
load("../data/african_gdps.RData")

tests <- 1000

options(scipen = 999)
```

We can visualize what years data for economies in African countries is given.

```{r countries-included, warning = F}

ggplot(data = african_gdps, mapping = aes(x = year)) +
  geom_histogram(binwidth = 10, fill = "blue", col = "black") +
  labs(
    x = "Year",
    y = "Number of Countries", 
    title = "Increase in Total Number of African
    Country Economic Observations Over Time"
    )
ggplot(data = african_gdps, mapping = aes(x = country, y = year)) +
  geom_point(size = .75, col = "orange") +
  coord_flip() +
  labs(
  x = "Country",
  y = "Year",
  title = "African Economic Data Observed Each Year by Country"
  )
```

We can also look at the distribution of GDP for observations since 2000 through visualization as a histogram and summary statistics.

```{r gdp-dist}
african_gdps %>%
  filter(!is.na(gdp)) %>%
  filter(year >= 2000) %>%
  ggplot(mapping = aes(x = gdp)) +
  geom_histogram(fill = "#5967EE", color = "black") +
  labs(title = "Distribution of African GDPs since 2000", x = "GDP (billions)", 
       y = "count")

african_gdps %>%
  filter(!is.na(gdp)) %>%
  filter(year >= 2000) %>%
  summarise(
    IQR = IQR(gdp),
    median = median(gdp)) %>%
  kable(col.names = c("Interquartile Range", "Median"), digits = 1)
```


## Data Analysis

## Question 1: How does independence impact a country's economic stability?

As African countries made a transition to becoming independent states, many were plagued by violence or political unrest. A bloody 10-year war in Algeria culminated in its independence; lengthy wars were fought in Angola and Mozambique, sectional divides fraught the Congo, and a political revolt ocurred in Kenya. Many of these movements brought economic instability to their countries. However, at the same time, these independence movements were fought in the hopes that establishing an independent nation would bring greater economic prosperity and stability to their country. 

Thus, on a broad level, we'd like to determine how these independence movements ultimately affected the stability of their countries' economies.  

Let's first do some exploratory data analysis. For each country, let's ask: how many years after independence will a country experience its next banking crisis?

```{r independence}
countries <- tibble(country = distinct(african_gdps, country)$country)
countries$independence_year <- african_gdps %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  arrange(year) %>%
  filter(row_number()==1) %>%
  ungroup() %>%
  select(year)
  
countries$crisis_year <- african_gdps %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  arrange(year) %>%
  filter(banking_crisis == "crisis") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(year)

countries <- countries %>%
  mutate(difference = (crisis_year$year - independence_year$year))

countries %>%
  ggplot(mapping = aes(x = reorder(country, difference), y =
                         difference)) +
  geom_bar(stat = "identity", fill = "blue") + 
  labs(title = "Years Between a Country 
    Achieving Independence and its Next Banking Crisis",
    y = "Number of Years", x = "Country") + 
  coord_flip()

countries %>%
  summarise(IQR = IQR(difference), median = median(difference), mean = mean(difference))

rm(countries)
```

We see that the median amount of years a country will first encounter a banking crisis after they achieve independence is about 28 years, with an interquartile range of 3 years and a mean of 31.3 years.

This is fascinating, but we're more interested in seeing if there's a difference in the economic stability of independent vs. colonized African countries. In particular, we're wondering if post-independence African countries see a higher proportion of systemic crises (per year) compared to before independence, when they were colonized. Let's examine it: 

```{r prop-crisis}
african_gdps %>%
  group_by(independence) %>%
  summarise(
    overall_crisis_prop = mean(systemic_crisis == 1)
  ) %>%
  mutate(independence = case_when(
    independence == 0 ~ "Independent Countries",
    independence == 1 ~ "Non-independent Countries"
  )) %>%
  kable(col.names = c("", "Proportion of Years with Crises"), digits = 4)
```

Based on our sample, we see that on average, there is a 0.42% chance of a systemic crisis occurring in any given year for a non-independent (i.e. colonized) country, while there is a 9.85% chance of a systemic crisis occurring in any given year for an independent country. The difference in these proportions is 0.0943; i.e. the chance of a systemic crises is 9.43 percentage points higher post-independence compared to pre-independence. 

We'd like to see if this relationship is statistically significant and holds for all African countries. We can do so by conducting a hypothesis test to see if the proportion of years with systemic crises for African countries is higher post-independence compared to pre-independence. Our null hypothesis is that the proportion of years with systemic crises between African countries post-independence and pre-independence is the same; the observed difference is due to chance. Our alternative hypothesis is that the proportion of years with systemic crises for African countries is higher post-independence than pre-independence.

Since we're testing for independence, we'll use permutation.

```{r crisis-prop-diff}
set.seed(1)
null_dist<- african_gdps %>%
  specify(response = systemic_crisis, explanatory = independence, 
          success = "1") %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in props", 
            order = c("1", "0"))
get_p_value(null_dist, obs_stat = 0.094320737, direction = "greater")

visualize(null_dist) +
  shade_p_value(0.094320737, "greater") +
  labs(
    title = "Null Distribution of Difference in Systemic Crisis Proportions",
    subtitle = "Between Independent and Non-Independent African Countries",
    x = "Difference in Proportions", 
    y = "Count")

rm(null_dist)
```

The p-value of our hypothesis test is 0. Since our p-value of 0 is less than our significance level of 0.05, we reject the null hypothesis. The data provides convincing evidence that the proportion of years with systemic crises for African countries is higher post-independence than pre-independence. 

Ultimately, we can conclude that after gaining independence, African countries experience greater economic instability compared to when they were under colonial rule.

## Question 2: Are there regional differences in economic development and stability between North African and sub-Saharan African Countries?

Today, many newspapers, academics, and policymakers classify Africa into two broad regions. There's North Africa, consisting of nations like Algeria and Morocco, and sub-Saharan Africa, which conists of countries ranging from the Central African Republic to Botswana. Traditionally, people associate sub-Saharan Africa with being less developed and more impoverished compared to the rest of the continent. In recent years, reports by the World Bank have claimed that more and more of the world's poor are being concentrated into a few sub-Saharan countries.

We set out to see if this is claim is true; that is, if there's a noticeable difference in economic stability and prosperity between North African and sub-Saharan African countries. We can evaluate these claims in two ways: by analyzing our data on each country's GDP and the proportion of years with systemic crises for each country.

#### Is the median GDP of North African countries greater than the median GDP of sub-Saharan countries?

To start, let's calculate the median GDP for North African and sub-Saharan countries. When calculating GDP by region, we'll use 2013 GDP data since it's recent and available for 11 of the 13 African countries in our dataset. 

```{r GDP-region-split, warning = F}
regional_med_gdps <- african_gdps %>%
  filter(!is.na(gdp)) %>%
  filter(year == 2013) %>%
  group_by(region) %>%
  summarise(med_gdp_billions = median(gdp))

rmg <- regional_med_gdps %>%
  pull(med_gdp_billions)


regional_med_gdps

rm(regional_med_gdps)
```

Let's also visualize the median GDP by region in 2013.

```{r visualize-median-GDP}
african_gdps %>%
  filter(year == 2013) %>%
  ggplot( mapping = aes(x = region, y = gdp)) +
  geom_boxplot(color = "black", fill = "#59CAEE") +
  labs(title = "Median 2013 GDP by Region", x = "Region", y = "Median GDP (Billion $ USD)")
```

From our visualization, we see that the interquartile range in median GDP for sub-Saharan countries is much larger than the interquartile range for North African countries, demonstrating larger variability. The country with the greatest GDP is a sub-Saharan country, which is an outlier for its region. However, the median 2013 GDP for sub-Saharan countries is less than the median 2013 GDP for North African countries.

The median GDP for North African countries is `r round(rmg[1], 1)` billion dollars; the median GDP for sub-Saharan countries is `r round(rmg[2], 1)` billion dollars. Therefore, the difference in median GDP between North African and sub-Saharan countries is `r round(abs(rmg[1] - rmg[2]), 1)` billion dollars. 

We'd like to see if this relationship is statistically significant and holds all North African and sub-Saharan countries. Thus, we'll ask the research question: is the median GDP of North African countries greater than the median GDP of sub-Saharan countries?

Our null hypothesis is that the median GDP of North African and sub-Saharan countries is the same; the observed difference is due to chance. Our alternative hypothesis is that the median GDP of North African countries is greater than the median GDP of sub-Saharan African countries.

Since we're testing for independence, we'll use permute.

```{r gdp-prop-diff}
set.seed(1)

null_dist <- african_gdps %>%
  filter(!is.na(gdp)) %>%
  filter(year == 2013) %>%
  specify(response = gdp, explanatory = region) %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in medians", 
            order = c("North Africa", "Sub-Saharan"))

visualize(null_dist) +
  shade_p_value(rmg[1] - rmg[2], "greater") +
  labs(title = "Null Distribution of Difference in 2013 Median GDP", subtitle = "Between North African and Sub-Saharan Countries", x = "Difference in GDP ($)", 
       y = "Count")

get_p_value(null_dist, obs_stat = rmg[1] - rmg[2], direction = "greater") %>% 
  kable(col.names = c("p-value"))

rm(null_dist)
rm(rmg)
```

The p-value of our hypothesis test is 0.339. Since our p-value of 0.339 is greater than our significance level of 0.05, we fail to reject the null hypothesis. The data does not provide convincing evidence that the median 2013 GDP of North African countries is greater than the median GDP of sub-Saharan countries.

#### Do sub-Saharan countries have a greater proportion of years with systemic crises than North African countries?

From our data sample, let's calculate the proportion of years with systemic crises for North African and sub-Saharan countries. 

```{r systemic-crisis-by-region}
african_gdps %>%
  group_by(region) %>%
  summarise(overall_crisis_prop = mean(systemic_crisis == 1)) %>%
  kable(col.names = c("Region", "Proportion of Years with Crisis"), digits = 4)
```

The proportion of years with systemic crises for North African countries is 0.0436; the proportion of years with systemic crises for sub-Saharan countries is 0.0971. The difference is 0.0536, i.e. the proportion of years with systemic crises is 5.36 percentage points higher for sub-Saharan countries compared to North African countries. 

We'd like to see if this relationship is statistically significant and holds for all North African and sub-Saharan countries. We can do so by conducting a hypothesis test to see if sub-Saharan countries have a greater proportion of years with systemic crises than North African countries.

Our null hypothesis is that the proportion of years with systemic crises between North African and sub-Saharan countries is the same; the observed difference is due to chance. Our alternative hypothesis is that the proportion of years with systemic crises for sub-Saharan countries is greater than for North African countries.

Since we're testing for independence, we'll use permute.

```{r region-prop-diff}
set.seed(1)

null_dist <- african_gdps %>%
  specify(response = systemic_crisis, explanatory = region, success = "1") %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in props", 
            order = c("Sub-Saharan", "North Africa"))

visualize(null_dist) +
  shade_p_value(0.0535702, "greater") +
  labs(title = "Null Distribution of Difference in Proportion of Systemic 
  Crises", subtitle = "Between North African and Sub-Saharan Countries", x = "Difference in Proportion", y = "Count")

get_p_value(null_dist, obs_stat = 0.0535702, direction = "greater") %>% 
  kable(col.names = c("p-value"))

rm(null_dist)

```

The p-value of our hypothesis test is 0. Since our p-value of 0 is less than the significance level of 0.05, we reject the null hypothesis. The data provides convincing evidence that sub-Saharan countries have a greater proportion of years with systemic crises compared to North African countries.

From our results, we can conclude that historically — when accounting for all years from the 19th century to today — sub-Saharan African countries have been more prone to systemic crises than North African countries. However, the data did not provide convincing evidence that the median 2013 GDP of North African countries was greater than the median 2013 GDP of sub-Saharan countries. That's a positive sign; it demonstrates that even though there have been historical disparities between these two regions, in the 21st century, sub-Saharan Africa is catching up, if not matching up to, North African economies.


## Question 3: Can we produce a model to predict economic crises in African economies?

```{r derivative-country-percentchange}

african_gdps <- african_gdps %>%
  group_by(country) %>%
  mutate(
    percentchange_gdp = (gdp - lag(gdp, default = gdp[1]))/gdp,
    percentchange_cpi = 
      (
        inflation_annual_cpi -
          lag(inflation_annual_cpi,default = inflation_annual_cpi[1])
      )
    /inflation_annual_cpi
  ) %>%
  filter(is.finite(percentchange_cpi))

```

To determine what factors most impacted African economies, we designed two linear regression models with the goal of using these to predict economic crises in African economies. Given the limits of our own statistical knowledge, we were not able to give a regression model that directly predicted a systemic crisis, though we used the nation’s percent change in CPI and percent change in GDP as indirect and discrete measures of a country’s state of crisis. We claimed that, if the percent change in GDP of a nation was particularly negative, or alternatively if the percent change in Consumer Price Index (CPI) – a measure used to study a nation’s inflation rate – was particularly high, we could claim that these indicated an economic crisis. The reasoning for using CPI and GDP as indicators of economic crisis is sound, according to external research into what can indicate economic crisis worldwide(Kuosmanen & Vataja 2013, Meyer & Zaman 2019), so we feel that our proxy metrics for predicting economic crisis are fairly strong. Even within our dataset, we can see that in crisis, median CPI is consistently higher, and median GDP is consistently lower.

```{r crisis-and-percentchange}
african_gdps %>%
  filter(!is.na(percentchange_gdp)) %>%
  group_by(systemic_crisis) %>%
  summarise(
    median_pcGDP = median(percentchange_gdp),
    iqr_pcGDP = IQR(percentchange_gdp),
    median_pcCPI = median(percentchange_cpi),
    iqr_pcCPI = IQR(percentchange_cpi)
  ) %>%
    kable(col.names = c(
      "Systemic Crisis",
      "Median %changeGDP",
      "IQR %changeGDP",
      "Median %changeCPI",
      "IQR %changeCPI"))
  
```

For each model, we examined each factor in the dataset and the two-factor interactions between every factor; we then performed backward selection on each model to determine the most useful variables for predicting their respective response variables (optimizing for AIC).

Below, we've given a breakdown of the two models: first, %change in GDP, and second, %change in CPI.

```{r full-model}

full_pcGDP_model <- lm(
  percentchange_gdp ~ 
    (year +
    gdp +
    exch_usd +
    domestic_debt_in_default +
    sovereign_external_debt_default +
    gdp_weighted_default +
    inflation_annual_cpi +
    percentchange_cpi +
    independence +
    region +
    currency_crises +
    inflation_crises +
    banking_crisis)^2,
  african_gdps)

full_pcCPI_model <- lm(
    percentchange_cpi ~ 
    (year +
    gdp +
    exch_usd +
    domestic_debt_in_default +
    sovereign_external_debt_default +
    gdp_weighted_default +
    inflation_annual_cpi +
    percentchange_gdp +
    independence +
    region +
    currency_crises +
    inflation_crises +
    banking_crisis)^2,
    african_gdps)
```

```{r step-func-AIC, results = "hide"}

best_aic_pcGDP <- step(full_pcGDP_model, direction = "backward")
best_aic_pcCPI <- step(full_pcCPI_model, direction = "backward")

rm(full_pcCPI_model, full_pcGDP_model)
```

```{r final-model}
tibble(c("%change GDP", "%change CPI"),c(glance(best_aic_pcGDP)$adj.r.squared, glance(best_aic_pcCPI)$adj.r.squared)) %>%
  kable(
    format = "markdown",
    digits = 3,
    col.names = c("", "Adjusted R-Squared"))


tidy(best_aic_pcGDP) %>%
  select(term, estimate, p.value) %>%
  arrange(p.value, abs(estimate)) %>%
  kable(format = "markdown", digits = 3)

tidy(best_aic_pcCPI) %>%
  select(term, estimate, p.value) %>%
  arrange(p.value, abs(estimate)) %>%
  kable(format = "markdown", digits = 3)

```

As these tables show, there are a lot of variables that go into predicting changes in GDP and CPI! We found several variables that fell within statistical significance (p < 0.05) for each variable, as can be seen above. To name a few interesting finds for each, we found that the percent change in GDP for African economies was highly negatively influenced, on average, by whether an inflation crisis was present or not. Likewise, we also found that, for Sub-Saharan nations, the debt-to-GDP ratio at the time of a default (`gdp_weighted_default`) predicted an average decrease in annual %change GDP, indicating a further crisis. Altogether, however, with an adjusted r-squared value of only about 23 percent, it is difficult to say whether this model can give accurate explanations for the variance in the data overall, and whether this model is truly useful for predicting annual %change GDP.

Conversely, our CPI model was shockingly accurate for such a complex metric, with an adjusted R-squared value of about 88 percent. That being said, we must note that the strange and large estimates the model provides - we found that they are largely the result of the inclusion of Zimbabwe, whose absolutely incredible currency and inflation crisis around 2009 onward has made percentages fall way above 100%. Nigeria also faced a far less extreme inflation crisis in 1972, which has also pushed annual %change in CPI far above 100%. While we contemplated making a model without these countries to even out the data, we felt that this omission would be to omit the realities of the data, extreme or not. Naturally, the exchange rate in USD and the presence of a currency crisis in an African economy was highly indicative of a positive annual %change CPI, on average; interestingly, though, defaulting on domestic debt without a banking crisis drove predicted a lower %change in CPI, on average. While we can't conclusively say why this is without further analysis, we might hypothesize that this is due to state amnesty - if a nation has to default on its domestic debt, the domestic banks responsible for the bailout might be incentivized to decrease the production of more currency to mitigate the losses of the debt.

Considering whether these models can be used for inference according to the Central Limit Theorem, we checked the four essential assumptions being made about the normality, spread, and independence of residuals and observations. Our residual graphs for each model are below.

```{r evaluator-func}
eval_assumptions <- function(model, lower_bound, upper_bound) {
  restrict_range_lower <- lower_bound
  restrict_range_upper <- upper_bound
  aug <- augment(model)
  
  indep_plot <- ggplot(data = aug, aes(x = 1:nrow(aug), y = .resid)) +
    geom_point(alpha = 0.4) +
    labs(title = "Are Observations Independent?", x = "Index", y = "Residual")
  
  distro_plot <- ggplot(data = aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, lty = 3, color = "gray") +
  labs(
    title = "Are Observations Randomly Distributed with Constant Variance?",
    y = "Residuals",
    x = "Predicted values, y-hat")
  
  normal_dist <- ggplot(data = aug, aes(x = .resid)) +
  geom_histogram(bins = 75) +
  labs(
    title = "Are Residuals Normally Distributed?",
    x = "Residuals")
  
  if (!missing(restrict_range_lower) && !missing(restrict_range_upper)) {
      
      q_resid <- quantile(aug$.resid,probs = c(restrict_range_lower, restrict_range_upper))
      
      q_fitted <- quantile(aug$.fitted, probs = c(restrict_range_lower, restrict_range_upper))
      
      indep_plot <- indep_plot + ylim(q_resid[1], q_resid[2])
      distro_plot <- distro_plot + xlim(q_fitted[1], q_fitted[2]) + ylim(q_resid[1], q_resid[2])
      normal_dist <- normal_dist + xlim(q_resid[1], q_resid[2])

  }
  name <- deparse(substitute(model)) %>%
    substring(., nchar(.) - 2, nchar(.))
  grid.arrange(
    top = paste("Model: Annual Percent Change in", name),
    indep_plot,
    distro_plot,
    normal_dist)
}
```

```{r assump_test_GDP, warning = F}
eval_assumptions(best_aic_pcGDP, 0.05, 0.95)
```

For our annual percent change in GDP regression model, it is clear that all conditions are met to use the model for inference. While it doesn't explain the variance of our dataset quite as well as our CPI model, it has randomly distributed residuals that are constantly variant and approximately normally distributed.

```{r assump_test_CPI, warning = F}
eval_assumptions(best_aic_pcCPI, 0.05, 0.95)
```

In contrast, while our annual percentage change in CPI regression model explains the variance in the dataset much better than our GDP model when comparing adjusted R-squared values, it may not be as suited to general inference. While the observations are independent and approximately normally distributed, there seems to be a clear trend in the distribution of observations. This indicates that while the model is quite precise in repeatedly predicting a similar value for any given factors, it is increasingly inaccurate in predicting larger changes, resulting in more residuals. Therefore, while this model is incredibly effective in predicting changes in CPI when CPI changes are small, it is not as accurate when CPI changes are large. For the purposes of our dataset, this is mostly fine - changes in CPI are generally incremental. In fact, the lack of data for larger changes in CPI can probably explain the inaccuracy of the model at these extremes.

## Conclusion

On the whole, we were able to better understand what affects systemic crises and GDP growth through our research project. Through hypothesis tests, we found that whether a country is independent or not greatly correlates with economic stability — after gaining independence, African countries experienced greater economic instability compared to when they were under colonial rule, measured by the proportion of years with systemic crises. To put this into statistical terms, we found that our observed difference in the proportion of years with systemic crisis between independent and non-independent countries would occur by chance less than 5% of the time (using a significance level of 0.05). Since our p-value was less than our significance level, we rejected the null hypothesis (that there was no difference in the proportion of years with systemic crises) and concluded that the data provides convincing evidence to support our alternative hypothesis.

We also found that regional differences, such as those between North Africa and sub-Saharan countries, play a role too. Although our hypothesis tests led us to conclude that there was no significant difference in the median 2013 GDP of each region, we did conclude that sub-Saharan African countries experienced a greater proportion of years with systemic crises compared to North African countries. This fact might be the most interesting of all: it demonstrates that even though there have been historical disparities between these two regions, in the 21st century, sub-Saharan Africa is catching up, if not matching up to, North African economies.

Our statistical analysis on economic stability used the proportion of years with systemic crises as a proxy for economic stability. While we think this is an accurate metric, our predictions could've been improved by using an actual metric of economic stability that combines many factors of economic health, similar to a former stability index produced by Lehman Brothers and Eurasia Group. 

An assumption of hypothesis tests is that the sample data is randomly sampled from the population (in this case, all African nations). However, our dataset, created for the purposes of economics research, definitely did not do perfectly random sampling to select the 13 African countries in the dataset. Perhaps these countries had a more thorough history of economic indicators, or perhaps they were more interesting to the compilers of the dataset. In any case, we note that this dataset is biased towards highly populated countries, like South Africa and Egypt. 

Moreover, our dataset was also inconsistent in other ways. Different countries were tracked in the data for different periods of time; data from South Africa started at 1900 while the first data from Nigeria occurs around 1955. There were also gaps in our data for some countries for a couple of years.

We also created two linear regression models to predict economic crises. While our models provide a fairly significant explanation for the variation in the dataset, with adjusted r-squared values of about 0.30 and 0.88 for annual %change in GDP and annual %change in CPI, respectively, it is difficult to confidently say whether they can predict economic crisis or not, given that we are not able to directly compare economic crisis and the explanatory factors of our models. In addition, our annual %change in CPI model, though precise (in the sense that residuals were consistently spread), was not very accurate at more extreme %changes in CPI, likely because of the lack of data provided for these extremes. That being said, given that most annual shifts in CPI are much more marginal than the wild changes seen in some African nations of late (notably Zimbabwe and Nigeria), the model does an excellent job during times of minor to moderate economic instability where %change in CPI is smaller year on year than catastrophic periods of rapid and immense change.

Finally, our model could certainly benefit from better generalizability. Given the constraints of our dataset and the scarcity of economic data for African countries prior to 1950, the amount of data provided for some African countries was so sparse that it would be difficult to confidently say that our model accurately explained their economic outcomes.

If we were to start over again, we would try to source a general dataset of economic information on all African countries, with data included for all the years we're interested in. This would make our hypothesis tests and linear regression model much more accurate, and give more accurate claims into whether there's truly differences between independence and non-independence. Of course, however, with the nature of historical economic data, that's hard to gather. If we were to continue on this project, we would be extremely interested in creating a logistic regression model to predict whether a systemic crisis will occur in any given year, given a set of economic indicators like inflation rate and exchange rate. Moreover, we'd also be interested in seeing if we could create a model that, given the years prior to a systemic crisis, can actually predict that a crisis will happen in the future. This would be a very powerful model, because it could help policymakers and business leaders understand that a crisis is looming, allowing them to prepare, mitigate, or solve these crises.


###### References