---
title: "An Analysis of African Systemic Crises"
author: "Beck Addison, Jerry Lin, Isabella Swigart, Emma Hirschkop"
date: "12/13/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

Your project writeup goes here! Before you submit, make sure your chunks are turned off with `echo = FALSE`. 

## Introduction
```{r load-packages, message = F}

#install.packages(c("infer","countrycode","pracma", "plotly"))
library(infer)
library(tidyverse)
library(countrycode)
library(pracma)
library(broom)
library(knitr)
library(plotly)
```

```{r load-data-and-initialization, message = F, warning = F}
africa <- read_csv("../data/african_crises.csv")
global <- read_csv("../data/global_crisis_data.csv")
gdps <- read_csv("../data/world_gdp_data.csv")

tests <- 1000

options(scipen = 999)
```

We can visualize what years data for economies in African countries is given.

```{r countries-included, warning = F}
ggplot(data = africa, mapping = aes(x = year)) +
  geom_histogram(binwidth = 10, fill = "blue", col = "black") +
  labs(
    x = "Year",
    y = "Number of Countries", 
    title = "Increase in Total Number of African
    Country Economic Observations Over Time"
    )
ggplot(data = africa, mapping = aes(x = country, y = year)) +
  geom_point(size = .75, col = "orange") +
  coord_flip() +
  labs(
  x = "Country",
  y = "Year",
  title = "African Economic Data Observed Each Year by Country"
  )
```


## Data Analysis

```{r data-cleaning, warning = F}
gdps <- gdps %>%
  pivot_longer(
    cols = c(
      -`Country Name`,
      -`Country Code`,
      -`Indicator Name`,
      -`Indicator Code`),
    names_to = "year"
  ) %>%
  rename(
    cc3 = `Country Code`,
    indicator_code = `Indicator Code`,
    indicator_name = `Indicator Name`,
    country = `Country Name`,
    gdp = value
    )

african_gdps <- gdps %>%
  merge(., africa, by = c("country","cc3","year"))



african_gdps <- african_gdps %>%
  mutate(
    year = as.numeric(year),
    systemic_crisis = factor(systemic_crisis),
    domestic_debt_in_default = factor(domestic_debt_in_default),
    sovereign_external_debt_default = factor(sovereign_external_debt_default),
    independence = factor(independence),
    currency_crises = factor(case_when(
      currency_crises < 1 ~ 0,
      TRUE ~ 1
    )),
    inflation_crises = factor(inflation_crises),
    banking_crisis = factor(banking_crisis)
  )

global <- global[-1,] %>%
  select(-`<`) %>%
  rename(
    case = Case,
    cc3 = CC3,
    country = Country,
    year = Year,
    banking_crisis = `Banking Crisis`,
    systemic_crisis = `Systemic Crisis`,
    gold_standard = `Gold Standard`,
    nat_currency = `national currency`,
    sov_ext_debt = 
    `SOVEREIGN EXTERNAL DEBT 1: DEFAULT and RESTRUCTURINGS, 1800-2012--Does not include defaults on WWI debt to United States and United Kingdom and post-1975 defaults on Official External Creditors`,
    sov_ext_debt_post1975 = `SOVEREIGN EXTERNAL DEBT 2: DEFAULT and RESTRUCTURINGS, 1800-2012--Does not include defaults on WWI debt to United States and United Kingdom but includes post-1975 defaults on Official External Creditors`,
    exc_primary = `exch_primary source code`,
    domest_debt_default = `Domestic_Debt_In_Default`,
    domest_debt_notes = `Domestic_Debt_ Notes/Sources`,
    default_ext_notes = `Defaults_External_Notes`,
    gdp_weighted_default = `GDP_Weighted_default`,
    inflat_aver_consumer_prices = `Inflation, Annual percentages of average consumer prices`,
    curr_crises = `Currency Crises`,
    independence = `Independence`,
    inflat_crises = `Inflation Crises`
) %>%
  mutate(
    banking_crisis = as.logical(as.numeric(
      banking_crisis
      )),
    systemic_crisis = as.logical(as.numeric(
      systemic_crisis
    )),
    gold_standard = as.logical(as.numeric(
      gold_standard
    )),
    domest_debt_default = as.logical(as.numeric(
      domest_debt_default
    )),
    independence = as.logical(as.numeric(
      independence
    )),
    curr_crises = as.logical(as.numeric(
      curr_crises
    )),
    inflat_crises = as.logical(as.numeric(
      inflat_crises
    ))
  )
```

## Question 1: How does independence impact a country's economic stability?

As African countries made a transition to becoming independent states, many were plagued by violence or political unrest. A bloody 10-year war in Algeria culminated in its independence; lengthy wars were fought in Angola and Mozambique, sectional divides fraught the Congo, and a political revolt ocurred in Kenya. Many of these movements brought economic instability to their countries. However, at the same time, these independence movements were fought in the hopes that establishing an independent nation would bring greater economic prosperity and stability to their country. 

Thus, on a broad level, we'd like to determine how these independence movements ultimately affected the stability of their countries' economies.  

Let's first do some exploratory data analysis. For each country, let's ask: how many years after independence will a country typically experience its next crisis?

```{r independence}
output <- tibble(country = distinct(africa, country)$country)
output$independence_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(year)
  
output$crisis_year <- africa %>%
  filter(independence == 1) %>%
  group_by(country) %>%
  filter(banking_crisis == "crisis") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(year)
output <- output %>%
  mutate(difference = (crisis_year$year - independence_year$year))
ggplot(data = output, mapping = aes(x = country, y = difference)) +
  geom_bar(stat = "identity", fill = "blue") + 
  labs(title = "Years Between a Country 
    Achieving Independence and Its Next Financial Crisis",
    y = "Number of Years", x = "Country") + 
  coord_flip()
output %>%
  summarise(
    IQR = IQR(difference),
    median = median(difference),
    mean = mean(difference)) %>%
  kable
```

We see that the median amount of years a country will first encounter a banking crisis after they achieve independence is about 30 years, with an interquartile range of 11 years and a mean of 31.3 years.

This is fascinating, but we're more interested in seeing if there's a difference in the economic stability of independent vs. colonized African countries. In particular, we're wondering if post-independence African countries see a higher proportion of systemic crises (per year) compared to before independence, when they were colonized. Let's examine it: 

```{r prop-crisis}
#per_country_prop <- africa %>%
 # group_by(country, independence) %>%
 # summarise(crisis_prop = sum(systemic_crisis)/n())
#ggplot(data=per_country_prop, aes(x=country, y=crisis_prop,fill=factor(independence))) + 
 # geom_bar(stat="identity", position=position_dodge()) +
 # coord_flip()
africa %>%
  group_by(independence) %>%
  summarise(overall_crisis_prop = sum(systemic_crisis)/n()) %>%
  mutate(independence = case_when(
    independence == 0 ~ "Independent Countries",
    independence == 1 ~ "Non-independent Countries"
  )) %>%
  kable(col.names = c("", "Proportion of Years with Crises"))
```

Based on our sample, we see that on average, there is a 0.42% chance of a systemic crisis occurring in any given year for a non-independent (i.e. colonized) country, while there is a 9.85% chance of a systemic crisis occurring in any given year for an independent country. The difference in these proportions is 0.094320737. 

We'd like to conduct a hypothesis test to see if the proportion of years with systemic crises for African countries is higher post-independence compared to pre-independence. Our null hypothesis is that the proportion of years with systemic crises between African countries post-independence and pre-independence is the same; the observed difference is due to chance. Our alternative hypothesis is that the proportion of years with systemic crises for African countries is higher post-independence than pre-independence.

Since we're testing for independence, we'll use permutation. We'll modify our dataset slightly by factoring success and systemic_crisis into a categorical variable so that it'll work nicely with infer. 

```{r systemic-categorical}
fct_africa <- africa %>%
  mutate(systemic_crisis = factor(systemic_crisis)) %>%
  mutate(independence = factor(independence))
```

```{r crisis-prop-diff}
set.seed(1)
null_dist <- fct_africa %>%
  specify(response = systemic_crisis, explanatory = independence, 
          success = "1") %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in props", 
            order = c("1", "0"))
get_p_value(null_dist, obs_stat = 0.094320737, direction = "greater") %>% kable
visualize(null_dist) +
  shade_p_value(0.094320737, "greater") +
  labs(
    title = "Null Distribution of Difference in Systemic Crisis Proportions",
    subtitle = "Between Independent and Non-Independent African Countries",
    x = "Difference in Proportions", 
    y = "Count"
    )
```

Since our p-value of 0 is less than our significance level of 0.05, we reject the null hypothesis. The data provides convincing evidence that the proportion of years with systemic crises for African countries is higher post-independence than pre-independence. 

Ultimately, we can conclude that after gaining independence, African countries experience greater economic instability compared to when they were under colonial rule.

## Question 2: Are there differences in economic development and stability between North African and sub-Saharan African Countries?

#### Is there a difference in GDP and frequency of systemic crises between North African and sub-Saharan African countries?

Today, many newspapers, academics, and policymakers classify Africa into two broad regions. There's North Africa, consisting of nations like Algeria and Morocco, and sub-Saharan Africa, which conists of countries ranging from the Central African Republic to Botswana. Traditionally, people associate sub-Saharan Africa with being less developed and more impoverished compared to the rest of the continent. In recent years, reports by the World Bank have claimed that more and more of the world's poor are being concentrated into a few sub-Saharan countries.

We set out to see if this is claim is true; that is, if there's a noticeable difference in economic stability and prosperity between North African and sub-Saharan African countries. We can evaluate these claims in two ways: by analyzing our data on each country's GDP and the proportion of years with systemic crises for each country.

To answer this question, we need to label North African and sub-Saharan countries in our Africa dataset. 

``` {r label-region}
fct_africa <- fct_africa %>%
  mutate(region = case_when(
    country == "Algeria" ~ "n",
    country == "Angola" ~ "s",
    country == "Central African Republic" ~ "s",
    country == "Egypt" ~ "n",
    country == "Ivory Coast" ~ "s",
    country == "Kenya" ~ "s",
    country == "Mauritius" ~ "s",
    country == "Morocco" ~ "n",
    country == "Nigeria" ~ "s",
    country == "South Africa" ~ "s",
    country == "Tunisia" ~ "n",
    country == "Zambia" ~ "s",
    country == "Zimbabwe" ~ "s"
  )) %>%
  mutate(region = case_when(
    region == "n" ~ "North Africa",
    region == "s" ~ "Sub-Saharan"
  ))
africa <- africa %>%
  mutate(region = case_when(
    country == "Algeria" ~ "n",
    country == "Angola" ~ "s",
    country == "Central African Republic" ~ "s",
    country == "Egypt" ~ "n",
    country == "Ivory Coast" ~ "s",
    country == "Kenya" ~ "s",
    country == "Mauritius" ~ "s",
    country == "Morocco" ~ "n",
    country == "Nigeria" ~ "s",
    country == "South Africa" ~ "s",
    country == "Tunisia" ~ "n",
    country == "Zambia" ~ "s",
    country == "Zimbabwe" ~ "s"
  )) %>%
  mutate(region = case_when(
    region == "n" ~ "North Africa",
    region == "s" ~ "Sub-Saharan"
  ))
african_gdps <- african_gdps %>%
  mutate(region = case_when(
    country == "Algeria" ~ "n",
    country == "Angola" ~ "s",
    country == "Central African Republic" ~ "s",
    country == "Egypt" ~ "n",
    country == "Ivory Coast" ~ "s",
    country == "Kenya" ~ "s",
    country == "Mauritius" ~ "s",
    country == "Morocco" ~ "n",
    country == "Nigeria" ~ "s",
    country == "South Africa" ~ "s",
    country == "Tunisia" ~ "n",
    country == "Zambia" ~ "s",
    country == "Zimbabwe" ~ "s"
  )) %>%
  mutate(region = case_when(
    region == "n" ~ "North Africa",
    region == "s" ~ "Sub-Saharan"
  ))
```

Let's calculate the median GDP for North African and sub-Saharan countries. When calculating GDP by region, we'll use 2013 GDP data since it's recent and available for 11 of the 13 African countries in our dataset. 

```{r GDP-region-split, warning = F}
regional_med_gdps <- african_gdps %>%
  filter(year == 2013) %>%
  group_by(region) %>%
  summarise(med_gdp = median(gdp))

regional_med_gdps %>% kable

rmg <- regional_med_gdps %>%
  pull(med_gdp)

african_gdps_2013 <- african_gdps %>%
  filter(year == 2013)

ggplot(data = african_gdps_2013, mapping = aes(x = region, y = gdp)) +
  geom_boxplot(color = "black", fill = "#59CAEE") +
  labs(title = "Median 2013 GDP by Region", x = "Region", y = "Median GDP ($)")

```

The IQR for sub-Saharan countries is much larger than the IQR for North African countries, demonstrating larger variability. The country with the greatest GDP is a sub-Saharan country, which is an outlier for its region. However, the median 2014 GDP for sub-Saharan countries is less than North African countries.

The median GDP for North African countries is `r rmg[1]`; the median GDP for sub-Saharan countries is `r rmg[2]`. Therefore, the difference in median GDP between North African and sub-Saharan countries is `r abs(rmg[1] - rmg[2])`.

The first research question we'll ask is: is the median GDP of North African countries greater than the median GDP of sub-Saharan countries?

Our null hypothesis is that the median GDP of North African and sub-Saharan countries is the same; the observed difference is due to chance. Our alternative hypothesis is that the median GDP of North African countries is greater than the median GDP of sub-Saharan African countries.

Since we're testing for independence, we'll use permute.
```{r gdp-prop-diff}
set.seed(1)
gdp_2013 <- african_gdps %>%
  filter(year == 2013)

null_dist2 <- gdp_2013 %>%
  specify(response = gdp, explanatory = region) %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in medians", 
            order = c("North Africa", "Sub-Saharan"))
get_p_value(null_dist2, obs_stat = rmg[1] - rmg[2], direction = "greater") %>% kable
visualize(null_dist2) +
  shade_p_value(rmg[1] - rmg[2], "greater") +
  labs(
    title = "Null Distribution of Difference in 2013 Median GDP",
    subtitle = "Between North African and Sub-Saharan Countries",
    x = "Difference in GDP ($)", 
    y = "Count")
```

Since our p-value of 0.339 is greater than our significance level of 0.05, we fail to reject the null hypothesis. The data does not provide convincing evidence that the median 2013 GDP of North African countries is greater than the median GDP of sub-Saharan countries.

Next, let's calculate the proportion of years with systemic crises for North African and sub-Saharan countries. 

```{r systemic-crisis-by-region}
africa %>%
  group_by(region) %>%
  summarise(overall_crisis_prop = sum(systemic_crisis)/n()) %>%
  kable
```

The proportion of years with systemic crises for North African countries is 0.0436; the proportion of years with systemic crises for sub-Saharan countries is 0.0971. The difference is 0.0535702.

The second research question we'll ask is: do sub-Saharan countries have a greater proportion of years with systemic crises than North African countries?

Our null hypothesis is that the proportion of years with systemic crises between North African and sub-Saharan countries is the same; the observed difference is due to chance. Our alternative hypothesis is that the proportion of years with systemic crises for sub-Saharan countries is greater than for North African countries.

Since we're testing for independence, we'll use permute.
```{r region-prop-diff}
set.seed(1)
null_dist3 <- fct_africa %>%
  specify(response = systemic_crisis, explanatory = region, success = "1") %>%
  hypothesize(null = "independence") %>%
  generate(tests, type = "permute") %>%
  calculate(stat = "diff in props", 
            order = c("Sub-Saharan", "North Africa"))
get_p_value(null_dist3, obs_stat = 0.0535702, direction = "greater") %>% kable
visualize(null_dist3) +
  shade_p_value(0.0535702, "greater") +
  labs(
  title = "Null Distribution of Difference in Proportion of Systemic 
  Crises",
  subtitle = "Between North African and Sub-Saharan Countries",
  x = "Difference in Proportion",
  y = "Count"
  )
```

Since our p-value of 0 is less than the significance level of 0.05, we reject the null hypothesis. The data provides convincing evidence that sub-Saharan countries have a greater proportion of years with systemic crises compared to North African countries.

From our results, we can conclude that historically — when accounting for all years from the 19th century to today — sub-Saharan African countries have been more prone to systemic crises than North African countries. However, the data did not provide convincing evidence that the median 2013 GDP of North African countries was greater than the median 2013 GDP of sub-Saharan countries. That's a positive sign; it demonstrates that even though there have been historical disparities between these two regions, in the 21st century, sub-Saharan Africa is catching up, if not matching up to, North African economies.


## Question 3: Can we produce a model to predict economic crises in African economies?


To determine what factors most impacted African economies, we designed three linear regression models with the goal of using these to predict economic crises in African economies. Given the limits of our own statistical knowledge, we were not able to give a regression model that directly predicted a systemic crisis, though we used the nation’s change in CPI (delta CPI) and change in GDP (delta GDP) as indirect and discrete measures of a country’s state of crisis. We claimed that, if the delta GDP of a nation was particularly negative, or if the delta Consumer Price Index (CPI) – a measure used to study a nation’s inflation rate – was particularly high, we could claim that these indicated an economic crisis.

After plotting the models and performing backwards selection, optimizing for AIC, we found that a significant (p < 0.001) negative correlation between a country in a currency crisis and whether they faced a systemic economic crisis overall; on average, a country undergoing a currency crisis was expected to have a delta GDP of about -4.74 billion USD, holding all else constant. Interestingly, we did not find a significant correlation (p = 0.073) between inflation crisis and delta GDP, nor did we find a significant correlation (p = 0.137) between a country’s debt-to-GDP ratio and delta GDP. Finally, we found that there was a significant positive correlation (p < 0.001) between a country’s GDP and changes to their GDP, though it was relatively minor – for every additional $USD in the country’s GDP, their delta GDP was expected to, on average, be about nine cents higher.


## Conclusion
Conclusion goes here.

### Critiques 

  While our models provide a fairly significant explanation for the variation in the dataset, with adjusted r-squared values of 0.3 or higher, it is difficult to confidently say whether they can predict economic crisis or not, given that we are not able to directly compare economic crisis and the explanatory factors of our models. In addition, our current models look at the change in GDP and CPI, which we believe to be a better indicator of economic stability than the GDP or CPI itself, we might improve our models by looking at the change in GDP as a proportion of the GDP overall. By doing this, we level our comparisons across countries such that a crisis for a small economy (where hundreds of millions USD might make the difference between stability and crisis) can be compared similarly to a crisis for a large economy (where a delta GDP of hundreds of billions USD might only make a small difference in the economic productivity of the nation).
  
  Finally, our model could certainly benefit from better generalizability. Given the constraints of our dataset and the scarcity of economic data for African countries prior to 1950, the amount of data provided for some African countries was so sparse that it would be difficult to confidently say that our model accurately explained their economic outcomes. 
